---
title: "Chapter 3 - Data and programming"
format: html
editor: visual
---

<https://book.stat420.org/data-and-programming.html>

# 3.1 Data types

-   Numeric

-   Integer

-   Complex

-   Logical

-   Character

# 3.2 Data structures

| Dimension | Homogeneous | Heterogeneous |
|-----------|-------------|---------------|
| 1         | Vector      | List          |
| 2         | Matrix      | Data Frame    |
| 3+        | Array       | \-            |

## 3.2.1 Vectors

```{r}
c(1, 3, 5, 7, 8, 9)
```

```{r}
x <- c(1, 3, 5, 7, 8, 9)
x
```

```{r}
c(42, "Statistics", TRUE)
```

```{r}
c(42, TRUE)
```

```{r}
(y = 1:100)
```

```{r}
# Scalar
2
```

```{r}
seq(from = 1.5, to = 4.2, by = 0.1)
```

```{r}
seq(1.5, 4.2, 0.1)
```

```{r}
rep("A", times = 10)
```

```{r}
rep(x, times = 3)
```

```{r}
c(x, rep(seq(1, 9, 2), 3), c(1, 2, 3), 42, 2:4)
```

```{r}
length(x)
length(y)
```

### 3.2.1.1 Subsetting

```{r}
x
```

```{r}
x[1]
```

```{r}
x[3]
```

```{r}
x[-2]
```

```{r}
x[1:3]
```

```{r}
x[c(1, 3, 4)]
```

```{r}
z = c(TRUE, TRUE, FALSE, TRUE, TRUE, FALSE)
z
```

```{r}
x[z]
```

## 3.2.2 Vectorization

```{r}
x <- 1:10
x + 1
2 * x
2 ^ x
sqrt(x)
log(x)
```

## 3.2.3 Logical operators

| Operator | Summary                          | Example                | Result |
|--------------|---------------------------|------------------|--------------|
| `x < y`  | `x` less than `y`                | `3 < 42`               | TRUE   |
| `x > y`  | `x` greater than `y`             | `3 > 42`               | FALSE  |
| `x <= y` | `x` less than or equal to `y`    | `3 <= 42`              | TRUE   |
| `x >= y` | `x` greater than or equal to `y` | `3 >= 42`              | FALSE  |
| `x == y` | `x` equal to `y`                 | `3 == 42`              | FALSE  |
| `x != y` | `x` not equal to `y`             | `3 != 42`              | TRUE   |
| `!x`     | not `x`                          | `!(3 > 42)`            | TRUE   |
| `x | y`  | `x` or `y`                       | `(3 > 42) | TRUE`      | TRUE   |
| `x & y`  | `x` and `y`                      | `(3 < 42) & (42 > 13)` | TRUE   |

```{r}
x <- c(1, 3, 5, 7, 8, 9)
x
```

```{r}
x > 3
```

```{r}
x < 3
```

```{r}
x == 3
```

```{r}
x != 3
```

```{r}
x == 3 & x != 3
```

```{r}
x == 3 | x != 3
```

```{r}
x[x > 3]
```

```{r}
x[x != 3]
```

```{r}
sum(x > 3)
```

```{r}
as.numeric(x > 3)
```

```{r}
which(x > 3)
```

```{r}
x[which(x > 3)]
```

```{r}
max(x)
```

```{r}
which(x == max(x))
```

```{r}
which.max(x)
```

## 3.2.4 More vectorization

```{r}
x <- c(1, 3, 5, 7, 8, 9)
y <- 1:100
```

```{r}
x + 2
```

```{r}
x + rep(2, 6)
```

```{r}
x > 3
```

```{r}
x > rep(3, 6)
```

```{r}
x + y
```

```{r}
length(x)
```

```{r}
length(y)
```

```{r}
length(y) / length(x)
```

```{r}
(x + y) - y
```

```{r}
y <- 1:60
x + y
```

```{r}
length(y) / length(x)
```

```{r}
rep(x, 10) + y
```

```{r}
all(x + y == rep(x, 10) + y)
```

```{r}
identical(x + y, rep(x, 10) + y)
```

```{r}
x <- c(1, 3, 5)
y <- c(1, 2, 4)
x == y
```

```{r}
all(x == y)
```

```{r}
any(x == y)
```

```{r}
x <- c(10 ^ (-8))
y <- c(10 ^ (-9))
```

```{r}
all(x == y)
```

```{r}
all.equal(x, y)
```

## 3.2.5 Matrices

```{r}
x <- 1:9
x
```

```{r}
X <- matrix(x, nrow = 3, ncol = 3)
X
```

```{r}
Y <- matrix(x, nrow = 3, ncol = 3, byrow = TRUE)
Y
```

```{r}
Z <- matrix(0, 2, 4)
Z
```

```{r}
X
X[1, 2]
X[, 2]
X[2, c(1, 3)]
```

```{r}
x <- 1:9
rev(x)
rep(1, 9)
```

```{r}
rbind(x, rev(x), rep(1, 9))
```

```{r}
cbind(col_1 = x, col_2 = rev(x), col_3 = rep(1, 9))
```

```{r}
x <- 1:9
y <- 9:1
X <- matrix(x, 3, 3)
Y <- matrix(y, 3, 3)
```

```{r}
X
Y
```

```{r}
X + Y
```

```{r}
# Element-wise multiplication
X * Y
```

```{r}
X / Y
```

```{r}
# Matrix multiplication
X %*% Y
```

```{r}
# Transpose of a matrix
t(X)
```

```{r}
Z <- matrix(c(9, 2, -3, 2, 4, -2, -3, -2, 16), 3, byrow = TRUE)
Z
```

```{r}
# Inverse of a matrix
solve(Z)
```

```{r}
# Verify that solve(Z) returns the inverse
solve(Z) %*% Z
```

```{r}
diag(3)
```

```{r}
all.equal(solve(Z) %*% Z, diag(3))
```

```{r}
X <- matrix(1:6, 2, 3)
X
```

```{r}
dim(X)
rowSums(X)
colSums(X)
rowMeans(X)
colMeans(X)
```

```{r}
Z
diag(Z)
```

```{r}
diag(1:5)
```

```{r}
diag(5)
```

**Calculations with vectors and matrices**

```{r}
a_vec <- c(1, 2, 3)
b_vec <- c(2, 2, 2)
```

```{r}
c(is.vector(a_vec), is.vector(b_vec))
```

```{r}
c(is.matrix(a_vec), is.matrix(b_vec))
```

```{r}
# inner product
a_vec %*% b_vec
```

```{r}
# outer product
a_vec %o% b_vec
```

When vectors are coerced to become matrices, they are column vectors.

```{r}
as.matrix(a_vec)
```

```{r}
as.matrix(a_vec) %*% b_vec
```

```{r}
# inner product
crossprod(a_vec, b_vec)
```

```{r}
# outer product
tcrossprod(a_vec, b_vec)
```

```{r}
C_mat = matrix(c(1, 2, 3, 4, 5, 6), 2, 3)
D_mat = matrix(c(2, 2, 2, 2, 2, 2), 2, 3)
```

```{r}
crossprod(C_mat, D_mat)
t(C_mat) %*% D_mat
```

```{r}
all.equal(crossprod(C_mat, D_mat), t(C_mat) %*% D_mat)
```

```{r}
crossprod(C_mat, C_mat)
t(C_mat) %*% C_mat
```

```{r}
all.equal(crossprod(C_mat, C_mat), t(C_mat) %*% C_mat)
```

## 3.2.6 Lists

```{r}
list(42, "Hello", TRUE)
```

```{r}
ex_list <- list(
  a = c(1, 2, 3, 4),
  b = TRUE,
  c = "Hello!",
  d = function(arg = 42) {print("Hello World!")},
  e = diag(5)
)

ex_list
```

Subset a list :

-   `$` returns a named element of a list

-   `[]` returns a list

-   `[[]]` returns an element of a list

```{r}
ex_list$e
```

```{r}
ex_list[1:2]
```

```{r}
ex_list[1]
```

```{r}
ex_list[[1]]
```

```{r}
ex_list[c("e", "a")]
```

```{r}
ex_list["e"]
```

```{r}
ex_list[["e"]]
```

```{r}
ex_list$d
```

```{r}
ex_list$d(arg = 1)
```

## 3.2.7 Data frames
