---
title: "Chapter 3 - Data and programming"
format: html
editor: visual
---

<https://book.stat420.org/data-and-programming.html>

# 3.1 Data types

-   Numeric

-   Integer

-   Complex

-   Logical

-   Character

# 3.2 Data structures

| Dimension | Homogeneous | Heterogeneous |
|-----------|-------------|---------------|
| 1         | Vector      | List          |
| 2         | Matrix      | Data Frame    |
| 3+        | Array       | \-            |

## 3.2.1 Vectors

```{r}
c(1, 3, 5, 7, 8, 9)
```

```{r}
x <- c(1, 3, 5, 7, 8, 9)
x
```

```{r}
c(42, "Statistics", TRUE)
```

```{r}
c(42, TRUE)
```

```{r}
(y = 1:100)
```

```{r}
# Scalar
2
```

```{r}
seq(from = 1.5, to = 4.2, by = 0.1)
```

```{r}
seq(1.5, 4.2, 0.1)
```

```{r}
rep("A", times = 10)
```

```{r}
rep(x, times = 3)
```

```{r}
c(x, rep(seq(1, 9, 2), 3), c(1, 2, 3), 42, 2:4)
```

```{r}
length(x)
length(y)
```

### 3.2.1.1 Subsetting

```{r}
x
```

```{r}
x[1]
```

```{r}
x[3]
```

```{r}
x[-2]
```

```{r}
x[1:3]
```

```{r}
x[c(1, 3, 4)]
```

```{r}
z = c(TRUE, TRUE, FALSE, TRUE, TRUE, FALSE)
z
```

```{r}
x[z]
```

## 3.2.2 Vectorization

```{r}
x <- 1:10
x + 1
2 * x
2 ^ x
sqrt(x)
log(x)
```

## 3.2.3 Logical operators

| Operator | Summary                          | Example                | Result |
|----------|----------------------------------|------------------------|--------|
| `x < y`  | `x` less than `y`                | `3 < 42`               | TRUE   |
| `x > y`  | `x` greater than `y`             | `3 > 42`               | FALSE  |
| `x <= y` | `x` less than or equal to `y`    | `3 <= 42`              | TRUE   |
| `x >= y` | `x` greater than or equal to `y` | `3 >= 42`              | FALSE  |
| `x == y` | `x` equal to `y`                 | `3 == 42`              | FALSE  |
| `x != y` | `x` not equal to `y`             | `3 != 42`              | TRUE   |
| `!x`     | not `x`                          | `!(3 > 42)`            | TRUE   |
| `x | y`  | `x` or `y`                       | `(3 > 42) | TRUE`      | TRUE   |
| `x & y`  | `x` and `y`                      | `(3 < 42) & (42 > 13)` | TRUE   |

```{r}
x <- c(1, 3, 5, 7, 8, 9)
x
```

```{r}
x > 3
```

```{r}
x < 3
```

```{r}
x == 3
```

```{r}
x != 3
```

```{r}
x == 3 & x != 3
```

```{r}
x == 3 | x != 3
```

```{r}
x[x > 3]
```

```{r}
x[x != 3]
```

```{r}
sum(x > 3)
```

```{r}
as.numeric(x > 3)
```

```{r}
which(x > 3)
```

```{r}
x[which(x > 3)]
```

```{r}
max(x)
```

```{r}
which(x == max(x))
```

```{r}
which.max(x)
```

## 3.2.4 More vectorization
