---
title: "Chapter 5 - Probability and statistics in R"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

<https://book.stat420.org/probability-and-statistics-in-r.html>

```{r}
#| include: false
library(tidyverse)
theme_set(theme_bw())
source("plot_functions.R")
```

```{r}
plot_pdf <- function(distr = "norm", mean = 0, sd = 1,
                     value) {
  
  ggplot() +
    geom_function(
      fun = dnorm, 
      args = list(mean = mean, sd = sd),
      linewidth = 0.5
      ) +
    geom_segment(aes(x = value, xend = value,
                     y = 0, 
                     yend = dnorm(x = value, mean = mean, sd = sd)),
               linetype = "dotted") + 
    geom_segment(aes(x = mean - (3 * sd), xend = value,
                     y = dnorm(x = value, mean = mean, sd = sd), 
                     yend = dnorm(x = value, mean = mean, sd = sd)),
               linetype = "dotted") +
    scale_x_continuous(
      limits = c(mean - (3 * sd), mean + (3 * sd)),
      breaks = seq(
        from = mean - (3 * sd),
        to = mean + (3 * sd),
        by = sd
        )
      ) +
    labs(x = "x", y = "f(x)",
         title = paste0("Normal distribution (mean = ", mean,
                        ", sd = ", sd, ")"),
         subtitle = paste0("PDF: x = ", value, 
                           ", f(x) = ", 
                           round(dnorm(x = value, mean = mean, sd = sd), 3))
    )
}

plot_cdf <- function(distr = "norm", mean = 0, sd = 1,
                     value_min = NULL, value_max = NULL) {
  
  if(is.null(value_min)) {
    value_min <- mean - 3 * sd
  }
  
  if(is.null(value_max)) {
    value_max <- mean + 3*sd
  }
  
  ggplot(data.frame(x = c(mean - 3*sd, mean + 3*sd)), aes(x)) +
    stat_function(fun = dnorm,
                  args = list(mean = mean, sd = sd)) +
    geom_area(stat = "function",
              fun = dnorm,
              args = list(mean = mean, sd = sd),
              fill = "#00998a",
              alpha = 0.5,
              xlim = c(value_min, value_max)) +
    scale_x_continuous(
      limits = c(mean - (3 * sd), mean + (3 * sd)),
      breaks = seq(
        from = mean - (3 * sd),
        to = mean + (3 * sd),
        by = sd
      )
    )
  
}
```

# 5.1 Probability in R

## 5.1.1 Distributions

-   `dname` calculates density (pdf) at input `x`

-   `pname` calculates distribution (cdf) at input `x`

-   `qname` calculates the quantile at in input probability

-   `rname` generates a random draw from a particular distribution

Consider a random variable $X$ which is $N(\mu=2, \sigma^2=25)$. To calculate the value of the pdf at `x = 3` (the height of the curve at `x = 3`):

```{r}
dnorm(x = 3, mean = 2, sd = 5)
```

```{r}
plot_norm_pdf(mu = 2, sigma = 5, value = 3)
```

To calculate the CDF at $x = 3$, that is $P(X \leq 3)$:

```{r}
pnorm(q = 3, mean = 2, sd = 5)
```

```{r}
plot_norm_cdf(mu = 2, sigma = 5, to = 3)
```

To calculate the quantile for probability $0.975$:

```{r}
qnorm(p = 0.975, mean = 2, sd = 5)
```

To generate a random sample of size $n=10$:

```{r}
rnorm(n = 10, mean = 2, sd = 5)
```

| Command  | Distribution |
|----------|--------------|
| `*binom` | Binomial     |
| `*t`     | t            |
| `*pois`  | Poisson      |
| `*f`     | F            |
| `*chisq` | Chi-Squared  |

```{r}
dbinom(x = 6, size = 10, prob = 0.75)
```

# 5.2 Hypothesis tests in R

## 5.2.1 One sample t-test: review

Suppose $x_i \sim  N(\mu, \sigma^2)$ and we want to test $H_0: \mu = \mu_0$ versus $H_1: \mu \neq \mu_0$.

Assuming $\sigma$ is unknown, we use the one-sample Student's $t$ test statistic:

$$
t=\frac{\bar{x}-\mu_0}{s/\sqrt{n}} \sim t_{n-1}
$$

Where

$$
\bar{x}=\frac{\sum_{i=1}^n x_i}{n}
$$

and

$$
s=\sqrt{\frac{1}{n-1} \sum_{i=1}^n(x_i-\bar{x})^2}
$$

A $100(1-\alpha)\%$ confidence interval for $\mu$ is given by:

$$
\bar{x} \pm t_{\alpha/2,n-1}\frac{s}{\sqrt{n}}
$$

Where $t_{\alpha/2,n-1}$ is the critical value such that $P(t > t_{\alpha/2,n-1})=\alpha/2$ for $n-1$ degrees of freedom.

## 5.2.2 One sample t-test: example

Suppose a grocery store sells "16 ounce" boxes of *Captain Crisp* cereal. A random sample of 9 boxes was taken and weighed. The weight in ounces is stored in data frame `capt_crisp`.

```{r}
capt_crisp = data.frame(weight = c(15.5, 16.2, 16.1, 15.8, 15.6, 16.0, 15.8, 15.9, 16.2))
```

```{r}
x_bar <- mean(capt_crisp$weight)
s <- sd(capt_crisp$weight)
mu_0 <- 16
n <- 9
```

Compute the test statistic:

```{r}
t <- (x_bar - mu_0) / (s / sqrt(n))
t
```

Compute the p-value of the test:

```{r}
pt(t, df = n - 1)
```

Using the built-in `t.test()` function:

```{r}
t.test(x = capt_crisp$weight, mu = 16, alternative = "less", conf.level = 0.95)
```

For a two-sided interval for the mean weight:

```{r}
capt_test_results <- t.test(capt_crisp$weight, mu = 16, alternative = "two.sided", conf.level = 0.95)
```

```{r}
names(capt_test_results)
```

```{r}
capt_test_results$conf.int
```

Compute the critical value for the confidence interval:

```{r}
qt(p = 0.975, df = 8)
```

Confidence interval:

```{r}
c(mean(capt_crisp$weight) - qt(0.975, df = 8) * sd(capt_crisp$weight) / sqrt(9), mean(capt_crisp$weight) + qt(0.975, df = 8) * sd(capt_crisp$weight) / sqrt(9))
```

## 5.2.3 Two sample t-test: review
